---
title: "IDS investigation worksheet"
author: "by Fire Emoji: Beth, Beatrice, Faaiz, Mohsin"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readxl)
library(RColorBrewer)
library(broom)
library(parsnip)
# Add any other libraries here



```


```{r load-data}
# load your data 
AVONET3_BirdTree <- read_excel("data/AVONET3_BirdTree.xlsx", 
    sheet = "AVONET3_BirdTree")

```

```{r}
Bird_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche) 
 



Total_Habitat <- Bird_data %>%
  group_by(Habitat) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  filter(Habitat != "NA")



ggplot(data  = Total_Habitat, 
       aes(x = Habitat,
           y = total.birds,
           fill = Habitat)) +
         geom_col() +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
      scale_fill_brewer(type = "div",
                        palette = 9,
                        direction = 1,
                        aesthetics = "fill") +
  labs(title = "Total bird count within each habitat",
       x = "Habitat",
       y = "Total Birds")
```


```{r}
forest_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche) %>%
  filter(Habitat == "Forest")

total_trophic <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  slice(c(-3))

forest_data <- left_join(total_trophic, Bird_data) %>%
  select(-Total.individuals)

green_shades <- c("#75975E", "#95BB72", "#B3CF99", "#DDEAD1")
green_shades <- green_shades[1:length(unique(Bird_data$Trophic.Niche))]


ggplot(data = total_trophic,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  scale_fill_manual(values = green_shades) +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Forest Habitat",
       x = "Trophic Level",
       y = "Number of Birds")
  

ggplot(data = forest_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = green_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each trophic level",
       subtitle = "Forest Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```
```{r}
shrubland_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Shrubland" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

shrubland_birds <- left_join(total_trophic_level, shrubland_birds, by = "Trophic.Level")
yellow_shades <- c("#FFFF4C","#FFFF90", "#FFFFC1", "#FFFFDD")
yellow_shades <- yellow_shades[1:length(unique(Bird_data$Trophic.Level))]

human_modified_birds <- human_modified_birds %>%
  filter(!is.na(Trophic.Level))
plot_data <- human_modified_birds %>%
  filter(!is.na(Trophic.Level), Trophic.Level != "NA")

ggplot(data = plot_data,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = yellow_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Shrubland Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = plot_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = yellow_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Shrubland Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}
marine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Marine" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

marine_birds <- left_join(total_trophic_level, marine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

purple_shades <- c("#ECE3FF", "#BFA9E0", "#A689E1")
purple_shades <- purple_shades[1:length(unique(Bird_data$Trophic.Level))]

purples <- c("#662D91","#A689E1","#D29EFF","#ECE3FF")

marine_birds <- marine_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = marine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = purples) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = marine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = purple_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Niche",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")

```
```{r}
#riverine birds data plot
#findings: no data available for scavenger culmen length, but there are <100 riverine scavenger birds, there is a significantly large amount of carnivore birds
#on average, the beak culmen of herbivores are greater

riverine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Riverine")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

riverine_birds <- left_join(total_trophic_level, riverine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

turqoise_shades <- c("#008081", "#81D8D0", "#B0DFE5")
turqoise_shades <- turqoise_shades[1:length(unique(Bird_data$Trophic.Level))]

turqoise <- c("#008081", "#81D8D0","#3FE0D0", "#B0DFE5")

riverine_birds <- riverine_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = riverine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = turqoise) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = riverine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = turqoise_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}
coastal_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Coastal")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

coastal_birds <- left_join(total_trophic_level, coastal_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

blue <- c("#0E4D92", "#4682B4","#73c2fb", "#89CFF0")
blue <- blue[1:length(unique(Bird_data$Trophic.Level))]

coastal_birds <- coastal_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = coastal_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = blue) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Coastal Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = coastal_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = blue) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Coastal Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}
#human modified birds data plot
#findings: there is <15 scavenger birds so barely shows up in the bar chart, but it shows that scavenger human modified birds have a greater beak culmen lengths on average compared to the other trophic levels
#couple of carnivore outliers

human_modified_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Human Modified" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

human_modified_birds <- left_join(total_trophic_level, human_modified_birds, by = "Trophic.Level")
brown_shades <- c("#5B3E31","#A08679", "#BCA89F", "#D8CBC4")
brown_shades <- brown_shades[1:length(unique(Bird_data$Trophic.Level))]

human_modified_birds <- human_modified_birds %>%
  filter(!is.na(Trophic.Level))
plot_data <- human_modified_birds %>%
  filter(!is.na(Trophic.Level), Trophic.Level != "NA")

ggplot(data = plot_data,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = brown_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Human Modified Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = plot_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = brown_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Human Modified Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}

# 5089-5093
#carnivoire <- Bird_data %>%
#filter(Trophic.Level == "Carnivore") %>%
  #arrange(Trophic.Niche) %>%
  #slice(- (5089:5093))
all_culmens <- Bird_data %>%
  filter(!is.na(Trophic.Level)) %>%
  arrange(Trophic.Level) %>%
filter(Trophic.Level != "NA")


  ggplot(data = all_culmens,
         mapping = aes(y = Beak.Length_Culmen, fill = Trophic.Level, x = Trophic.Level)) +
  geom_boxplot() +
    scale_fill_manual( values = c(
      "Carnivore" = "#f0f9e8",
      "Herbivore" = "#bae4bc",
      "Omnivore" = "#7bccc4",
      "Scavenger" = "#2b8cbe"
    )) +
    labs(title = "Culmen Lengths",
         subtitle = "by Trophic Level",
         y = "Culmen Length (cm)",
         x = "Trophic Level",
         fill = "Trophic Level") 
  
```

```{r}
carnivoire <- Bird_data %>%
filter(Trophic.Level == "Carnivore") %>%
  arrange(Trophic.Niche) %>%
  filter(Trophic.Niche != "NA")
```



```{r}
birds.data_mass <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Mass, Trophic.Level)%>%
  mutate(Mass_kg = Mass/1000,
         log_mass_kg = log(Mass_kg), log_culmen = log(Beak.Length_Culmen)) %>%
   filter(Trophic.Level != "NA")

ggplot(data = birds.data_mass,
       mapping = aes(x = log_culmen,
                     y = log_mass_kg)) +
  geom_point(alpha = 0.3, mapping = aes(colour = Trophic.Level)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    x = "Log of Culmen Length (cm)", 
    y = "Log of Mass (kg)",
    fill = "Trophic Level"
  ) 
# 1% increase in x = 2.198% in y
```
```{r}

linear_reg() %>%
  set_engine("lm") %>%
  fit(log_mass_kg ~ log_culmen, data = birds.data_mass) %>%
  tidy()
```