---
title: "IDS investigation worksheet"
author: "by Fire Emoji: Beth, Beatrice, Faaiz, Mohsin"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readxl)
library(RColorBrewer)
# Add any other libraries here



```


```{r load-data}
# load your data 
AVONET3_BirdTree <- read_excel("data/AVONET3_BirdTree.xlsx", 
    sheet = "AVONET3_BirdTree")
View(AVONET3_BirdTree)


```

```{r}
Bird_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche) 
 



Total_Habitat <- Bird_data %>%
  group_by(Habitat) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  slice(- 7)



ggplot(data  = Total_Habitat, 
       aes(x = Habitat,
           y = total.birds,
           fill = Habitat)) +
         geom_col() +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
      scale_fill_brewer(type = "div",
                        palette = 9,
                        direction = 1,
                        aesthetics = "fill") +
  labs(title = "Total bird count within each habitat",
       x = "Habitat",
       y = "Total Birds")
```


```{r}
forest_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche) %>%
  filter(Habitat == "Forest")

total_trophic <- AVONET3_BirdTree %>%
  group_by(Trophic.Niche) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  slice(c(1,2,3,4,5,6,8,9,10,11))

forest_data <- left_join(total_trophic, Bird_data) %>%
  select(-Total.individuals)

pink_shades <- c("#C21E56", "#EC008C", "#FD3F92", "#F9429E", "#FF69B4",
                 "#FD6C9E", "#FC89AC", "#F6ADC6", "#FFC1CC", "#FDDDE6")
pink_shades <- pink_shades[1:length(unique(Bird_data$Trophic.Niche))]


ggplot(data = total_trophic,
       mapping = aes(x = Trophic.Niche,
                     y = total.birds,
                     fill = Trophic.Niche)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

ggplot(data = forest_data,
       mapping = aes( x = Trophic.Niche,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Niche)) +
  geom_boxplot() +
  scale_fill_manual(values = pink_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each trophic niche",
       subtitle = "Forest Habitat",
       x = "Trophic Niche",
       y= "Beak Length Culmen")
```


```{r}
marine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Marine" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

marine_birds <- left_join(total_trophic_level, marine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

purple_shades <- c("#ECE3FF", "#BFA9E0", "#A689E1")
purple_shades <- purple_shades[1:length(unique(Bird_data$Trophic.Niche))]

purples <- c("#662D91","#A689E1","#D29EFF","#ECE3FF")

marine_birds <- marine_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = marine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = purples) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = marine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = purple_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Niche",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")

```
```{r}
riverine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Riverine")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

riverine_birds <- left_join(total_trophic_level, riverine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

turqoise_shades <- c("#008081", "#81D8D0", "#B0DFE5")
turqoise_shades <- turqoise_shades[1:length(unique(Bird_data$Trophic.Niche))]

turqoise <- c("#008081", "#81D8D0","#3FE0D0", "#B0DFE5")

riverine_birds <- riverine_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = riverine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = turqoise) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = riverine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = turqoise_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```
```{r}
coastal_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Coastal")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

coastal_birds <- left_join(total_trophic_level, coastal_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(!is.na(Trophic.Level), total.birds > 0)

blue_shades <- c("#0E4D92", "#4682B4","#89CFF0")
blue_shades <- blue_shades[1:length(unique(Bird_data$Trophic.Niche))]

blue <- c("#0E4D92", "#4682B4","#73c2fb", "#89CFF0")

coastal_birds <- coastal_birds %>%
  filter(!is.na(Trophic.Level))

ggplot(data = coastal_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = blue) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Coastal Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = coastal_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = blue_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Coastal Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```



```{r}

# 5089-5093
carnivoire <- Bird_data %>%
filter(Trophic.Level == "Carnivore") %>%
  arrange(Trophic.Niche) %>%
  slice(- (5089:5093))

  ggplot(data = carnivoire,
         mapping = aes(y = Beak.Length_Culmen, fill = Trophic.Niche, x = Trophic.Niche)) +
  geom_boxplot() +
    scale_fill_manual( values = c(
      "Aquatic predator" = "#f0f9e8",
      "Invertivore" = "#bae4bc",
      "Omnivore" = "#7bccc4",
      "Vertivore" = "#2b8cbe"
    )) +
    labs(title = "Carnivore Culmen Lengths",
         subtitle = "by Trophic Niche",
         y = "Culmen Length (cm)",
         x = "Trophic Niche",
         fill = "Trophic Niche") 
  
```




  
  
```{r}
carnivoire <- Bird_data %>%
filter(Trophic.Level == "Carnivore") %>%
  arrange(Trophic.Niche) %>%
  slice(- (5089:5093))
```



```{r}
birds.data_mass <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Mass)%>%
  mutate(Mass_kg = Mass/1000)

ggplot(data = birds.data_mass,
       mapping = aes(x = Beak.Length_Culmen,
                     y = Mass_kg)) +
  geom_point()

```

