---
title: "IDS investigation worksheet"
author: "by Fire Emoji: Beth, Beatrice, Faaiz, Mohsin"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readxl)
library(RColorBrewer)
library(broom)
library(parsnip)
library(tidymodels)
library(knitr)
library(dplyr)

# Add any other libraries here



```


```{r load-data}
# load your data 
AVONET3_BirdTree <- read_excel("data/AVONET3_BirdTree.xlsx", 
    sheet = "AVONET3_BirdTree")

```


```{r}
Bird_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche, Mass) %>%
  filter(Total.individuals > 5)
 

Total_Habitat <- Bird_data %>%
  group_by(Habitat) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  filter(Habitat != "NA")


ggplot(data  = Total_Habitat, 
       aes(x = Habitat,
           y = total.birds,
           fill = Habitat)) +
         geom_col() +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
      scale_fill_brewer(type = "div",
                        palette = 9,
                        direction = 1,
                        aesthetics = "fill") +
  labs(title = "Total bird count within each habitat",
       x = "Habitat",
       y = "Total Birds")

```

```{r}
expl.stats_birds <- Bird_data %>%
  group_by(Trophic.Level) %>%
  summarise(avg_culmen.length = mean(Beak.Length_Culmen,),
         avg_mass = mean(Mass)) %>%
  select(Trophic.Level, avg_culmen.length, avg_mass)

```


```{r}


total_trophic <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  filter(Trophic.Level != "NA")

pink <- c("#FFC4D6", "#FFA6C1", "#FF87AB", "#FF5D8F")

ggplot(data = total_trophic,
       aes(x = Trophic.Level,
           y = total.birds,
           fill = Trophic.Level)) +
  geom_col() +
  theme_minimal() +
  scale_fill_manual(values = pink) +
  labs(title = "Total bird count within each Trophic Level",
       x = "Trophic Level",
       y = "Total Birds")

```


```{r}
#forest birds data plot

forest_data <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level, Trophic.Niche) %>%
  filter(Habitat == "Forest")

total_trophic <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) %>%
  slice(c(-3))

forest_data <- left_join(total_trophic, Bird_data) %>%
  select(-Total.individuals)

green_shades <- c("#75975E", "#95BB72", "#B3CF99", "#DDEAD1")
green_shades <- green_shades[1:length(unique(Bird_data$Trophic.Niche))]


ggplot(data = total_trophic,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  scale_fill_manual(values = green_shades) +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Forest Habitat",
       x = "Trophic Level",
       y = "Number of Birds")
  

ggplot(data = forest_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = green_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each trophic level",
       subtitle = "Forest Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}
#grassland bird data plot

grassland_birds <- AVONET3_BirdTree %>%
  select(Species3, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Grassland")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

grassland_birds <- left_join(total_trophic_level, grassland_birds, by = "Trophic.Level") %>%
  filter(Trophic.Level != "NA", total.birds > 0)

greens <- c("#57B000", "#63CA00","#70E300", "#7CFC00")
greens <- greens[1:length(unique(Bird_data$Trophic.Level))]

grassland_birds <- grassland_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = grassland_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = greens) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Grassland Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = grassland_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = greens) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Grassland Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}
#woodland bird data plot

woodland_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Woodland")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

woodland_birds <- left_join(total_trophic_level, woodland_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

brown <- c("#542A0C","#7E3F12","#934A15", "#BD5F1B")
brown <- brown[1:length(unique(Bird_data$Trophic.Level))]

woodland_birds <- woodland_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = woodland_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = brown) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Woodland Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = woodland_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = brown) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Woodland Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}
#wetland birds data plot

wetland_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Wetland")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

wetland_birds <- left_join(total_trophic_level, wetland_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

lightblue <- c("#6FC0DB","#ADD8E6","#C3E0EA", "#E1EEF2")
lightblue <- lightblue[1:length(unique(Bird_data$Trophic.Level))]

wetland_birds <- wetland_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = wetland_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = lightblue) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Wetland Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = wetland_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = lightblue) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Woodland Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}
# shrubland birds data plot

shrubland_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Shrubland" & (Trophic.Level == "Scavenger" | Trophic.Level != "NA"))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

shrubland_birds <- left_join(total_trophic_level, shrubland_birds, by = "Trophic.Level")
yellow_shades <- c("#FEBE00","#FDDB53", "#F8E472", "#FEFBBD")
yellow_shades <- yellow_shades[1:length(unique(Bird_data$Trophic.Level))]

shrubland_birds <- shrubland_birds %>%
  filter(Trophic.Level != "NA")
plots_data <- shrubland_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = plots_data,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = yellow_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Shrubland Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = plots_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = yellow_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Shrubland Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}
#desert birds data plot

desert_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Desert" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

desert_birds <- left_join(total_trophic_level, desert_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

orange_shades <- c("#FF6700", "#FF8C00", "#FF9F00", "#FFBF00")
orange_shades <- orange_shades[1:length(unique(Bird_data$Trophic.Level))]

desert_birds <- desert_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = desert_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = orange_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Desert Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = desert_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = orange_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Niche",
       subtitle = "Desert Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")

```

```{r}
#rock birds data plot

rock_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Rock" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

rock_birds <- left_join(total_trophic_level, rock_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

grey_shades <- c("#515151", "#737373", "#959595", "#B6B6B6")
grey_shades <- grey_shades[1:length(unique(Bird_data$Trophic.Level))]

rock_birds <- rock_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = rock_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = grey_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Rock Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = rock_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = grey_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Niche",
       subtitle = "Rock Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")

```


```{r}
#marine birds data plot

marine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Marine" & (Trophic.Level == "Scavenger" | !is.na(Trophic.Level)))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

marine_birds <- left_join(total_trophic_level, marine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

purple_shades <- c("#ECE3FF", "#BFA9E0", "#A689E1")
purple_shades <- purple_shades[1:length(unique(Bird_data$Trophic.Level))]

purples <- c("#662D91","#A689E1","#D29EFF","#ECE3FF")

marine_birds <- marine_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = marine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = purples) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = marine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = purple_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Niche",
       subtitle = "Marine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")

```


```{r}
#riverine birds data plot

#findings: no data available for scavenger culmen length, but there are <100 riverine scavenger birds, there is a significantly large amount of carnivore birds
#on average, the beak culmen of herbivores are greater

riverine_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Riverine")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

riverine_birds <- left_join(total_trophic_level, riverine_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

turqoise_shades <- c("#008081", "#81D8D0", "#B0DFE5")
turqoise_shades <- turqoise_shades[1:length(unique(Bird_data$Trophic.Level))]

turqoise <- c("#008081", "#81D8D0","#3FE0D0", "#B0DFE5")

riverine_birds <- riverine_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = riverine_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = turqoise) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = riverine_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = turqoise_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Riverine Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}
#coastal birds data plot

coastal_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Coastal")

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

coastal_birds <- left_join(total_trophic_level, coastal_birds, by = "Trophic.Level") %>%
  select(-Total.individuals) %>%
  filter(Trophic.Level != "NA", total.birds > 0)

blue <- c("#0E4D92", "#4682B4","#73c2fb", "#89CFF0")
blue <- blue[1:length(unique(Bird_data$Trophic.Level))]

coastal_birds <- coastal_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = coastal_birds,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = blue) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = " Coastal Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = coastal_birds,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = blue) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Coastal Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```


```{r}
#human modified birds data plot

#findings: there is <15 scavenger birds so barely shows up in the bar chart, but it shows that scavenger human modified birds have a greater beak culmen lengths on average compared to the other trophic levels
#couple of carnivore outliers

human_modified_birds <- AVONET3_BirdTree %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Trophic.Level) %>%
   filter(Habitat == "Human Modified" & (Trophic.Level == "Scavenger" | Trophic.Level != "NA"))

total_trophic_level <- AVONET3_BirdTree %>%
  group_by(Trophic.Level) %>%
  summarise(total.birds = sum(Total.individuals)) 

human_modified_birds <- left_join(total_trophic_level, human_modified_birds, by = "Trophic.Level")
brown_shades <- c("#5B3E31","#A08679", "#BCA89F", "#D8CBC4")
brown_shades <- brown_shades[1:length(unique(Bird_data$Trophic.Level))]

human_modified_birds <- human_modified_birds %>%
  filter(Trophic.Level != "NA")
plot_data <- human_modified_birds %>%
  filter(Trophic.Level != "NA")

ggplot(data = plot_data,
       mapping = aes(x = Trophic.Level,
                     y = total.birds,
                     fill = Trophic.Level)) +
  geom_col() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  scale_fill_manual(values = brown_shades) +
  labs(title = "Number of Birds in each Trophic Level",
       subtitle = "Human Modified Habitat",
       x = "Trophic Level",
       y = "Number of Birds")

ggplot(data = plot_data,
       mapping = aes( x = Trophic.Level,
                      y = Beak.Length_Culmen,
                      fill = Trophic.Level)) +
  geom_boxplot() +
  scale_fill_manual(values = brown_shades) +
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) +
  labs(title = "Culmen length of each Trophic Level",
       subtitle = "Human Modified Habitat",
       x = "Trophic Level",
       y= "Beak Length Culmen")
```

```{r}

#box plots of all the habitats' beak culmen length against trophic levels

mean_culmen <- Bird_data %>%
  group_by(Habitat, Trophic.Level) %>%
  summarise(mean_culmen = mean(Beak.Length_Culmen, na.rm = TRUE)) %>%
  ungroup()

blue <- c("#0E4D92", "#4682B4","#73c2fb", "#89CFF0")
blue <- blue[1:length(unique(Bird_data$Trophic.Level))]

clean_data <- Bird_data %>%
  filter(
    !is.na(Habitat),
    !is.na(Trophic.Level),
    Trophic.Level != "NA",
    Habitat != "NA"
  )

ggplot(
  data= clean_data %>%
    filter(!is.na(Habitat), !is.na(Trophic.Level)),
  aes(x = Trophic.Level,
      y = Beak.Length_Culmen,
      fill = Trophic.Level)
  ) +
  
  geom_boxplot(outlier.size = 0.8) +
  facet_wrap(~ Habitat, ncol = 3, scales = "free_y") +
  scale_y_log10() +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(size = 7, angle = 20),
        strip.text = element_text(size = 10, face = "bold"),
        legend.position = "bottom") +
  scale_fill_manual(values = blue) +
  labs(
    title = "Culmen Beak Length across different Trophic Levels",
    subtitle = "Seperated by Habitat",
    x = "Trophic Level",
    y = "Log of the Culmen Beak Length"
  )
```

```{r}
birds.data_mass <- Bird_data %>%
  select(Species3, Total.individuals, Beak.Length_Culmen, Habitat, Mass, Trophic.Level)%>%
  mutate(Mass_kg = Mass/1000,
         log_mass_kg = log(Mass_kg), log_culmen = log(Beak.Length_Culmen)) %>%
   filter(Trophic.Level != "NA")

ggplot(data = birds.data_mass,
       mapping = aes(x = log_culmen,
                     y = log_mass_kg)) +
  geom_point(alpha = 0.6, size = 0.6, mapping = aes(colour = Trophic.Level)) +
  geom_smooth(method = lm, se = FALSE, formula = y ~ x) +
  labs(
    title = "Bird Culmen Length vs. Mass",
    subtitle = "by trophic level",
    x = "Log of Culmen Length (mm)", 
    y = "Log of Mass (kg)",
    colour = "Trophic Level"
  ) +
  scale_colour_manual(values = c(
    "Carnivore" = "#88419d",
    "Herbivore" = "#d7b5d8",
    "Omnivore" = "#41b6c4",
    "Scavenger" = "#0570b0"
  )) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12))
# 1% increase in x = 2.198% in y
```
```{r}

linear_culmen_mass <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log_mass_kg ~ log_culmen, data = birds.data_mass)

tidy(linear_culmen_mass)

glance(linear_culmen_mass)
```

```{r}

ggplot(Bird_data, 
       aes(
         x = Beak.Length_Culmen, 
         weight = Total.individuals)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue") +
  labs(title = "Histogram of Beak Lengths (Weighted by Individuals)",
       x = "Beak Length (Culmen)", 
       y = "Total Individuals") +
  scale_x_continuous(limits = c(0, 100)) +
  theme_minimal() +
  geom_vline(xintercept = c(8, 25), 
             linetype = "dashed", 
             color = "red")  
# scale_x_log10()

# beak length values exist beyond 250, but for the sake of simplicity the graph is restricted to values uptil 100.

Bird_data %>%
  arrange(Total.individuals)

```

```{r}
#bird_fit <- linear_reg() %>%
  #set_engine("lm") %>%
  #fit(Beak.Length_Culmen ~ Habitat + Trophic.Level + Mass, data = train_data) %>%
  #tidy()

#bird_fit


size_data <- mutate(Bird_data,
                    size =
                    factor(case_when(
                      Beak.Length_Culmen < 20.0 ~ "small",
                      Beak.Length_Culmen >= 20.0 ~ "big"
                      )))

set.seed(123)

bird_split <- initial_split(size_data, prop = 0.80)

train_data <- training(bird_split)
test_data  <- testing(bird_split)

bird_rec <- recipe(
  size ~ Habitat + Trophic.Level + Mass,       
  data = train_data 
  ) %>%
  step_dummy(all_nominal(), -all_outcomes())

summary(bird_rec)

bird_mod <- logistic_reg() %>%
  set_engine("glm")


bird_wkflow <- workflow() %>% 
  add_model(bird_mod) %>% 
  add_recipe(bird_rec)

bird_fit <- bird_wkflow %>%
  fit(data = train_data)

bird_pred <- predict(bird_fit, test_data, type = "prob") %>% 
  bind_cols(test_data)

bird_pred

bird_pred %>%
  roc_curve(
    truth = size,
    .pred_small,
    event_level = "second"
  ) %>%
  autoplot()


bird_pred %>%
  roc_auc(
    truth = size,
    .pred_small,
    event_level = "second"
  )

cutoff_prob <- 0.5
bird_table <- bird_pred %>%
  mutate(
    size      = if_else(size == "small", "Bird beak is small", "Bird beak is big"),
    size_pred = if_else(.pred_small > cutoff_prob, "Bird beak labelled as small", "Bird beak labelled as big")
    ) %>%
  count(size_pred, size) %>%
  pivot_wider(names_from = size, values_from = n) 
```


```{r}
bird_table
#%>%
#  kable(col.names = c("", "Bird Beak is big", "Bird Beak is small"))

sensitivity <- 439/(439+33)
sensitivity
specificity <- 285/(285+137)
specificity
```